Pragmatics account for apparent limits on theory of mind use in adults

### 2016-01-27 writing/cogsci-submission/cogsci_keysar.tex
Fill out discussion

\begin{figure*}[t!]
\begin{center}
\includegraphics[scale = .35]{images/overall_view.jpg}
\end{center}
\vspace{-.25cm}
\caption{Interface used in the reported experiments. Objects behind the black squares were hidden from the director. \todo[inline]{TODO: Use images from tape condition, add extra panel to show matcher view.}}
\label{fig:interface}
\end{figure*}

@@

In this paper we argue that the apparent failures this view is based on are in fact successes for sophisticated social reasoning. 
In particular, we argue that critical utterances used by Keysar and colleagues are \emph{uncooperative}: they are less informative than what a speaker would actually produce in that situation; listeners expect these more informative utterances, and interpret the actual utterances accordingly---resulting in ``errors''.

The argument of Keysar and colleagues is based on an elegant experimental paradigm, where participants played a simple communication game with a confederate. The two players were placed on opposite sides of a $4 \times 4$ grid containing a set of everyday objects.
\todo[inline]{Refer to updated Fig.~1 with both director screen and matcher screen }
The confederate played the role of `director,' giving instructions about how to move objects around a grid, and the participant played the role of `matcher,' attempting to follow these instructions. For example, the objects in one trial included a medium-sized measuring cup in one location and a smaller measuring cup in another location. The director gave an instruction like `move the large cup up one square,' referring to the medium-sized one. 
Some objects were occluded such that only the matcher could see them, creating an asymmetry in the players' knowledge. To perform accurately on critical trials, the matcher would need to apply theory of mind to reason about the director's beliefs. For example, imagine a much larger measuring cup were placed in an occluded slot: if a participant failed to account for the director's (partial) knowledge, she might interpret `the large cup' to mean this occluded cup (which the director couldn't possibly know about). 
Indeed, \citeA{KeysarLinBarr03_LimitsOnTheoryOfMindUse} found that participants attempted to move the hidden item in 30\% of cases: 71\% of participants attempted to move this hidden item at least once (out of four critical cases) in the experimental condition, compared to 0\% in a control condition where there was no ambiguity over the referent. 
%46\% of participants reached for the hidden object at least twice. 
Additionally, eye-tracking data showed that participants considered the hidden item more often and for longer in the experimental condition than the control condition.

While these results are compelling, the paradigm has been criticized from several different angles. First, in all experiments, the privileged object was a better fit for the referring expression than the one in common ground (e.g. the largest measuring cup vs. the medium measuring cup for ``the large measuring cup"), making the privileged object \emph{a priori} more likely to be the referent. It would be fairer to compare two objects that fit the referring expression equally well \cite{HellerGrodnerTanenhaus08_Perspective}. Second, the viewpoint asymmetry paradigm is somewhat unnatural: common ground is typically built incrementally over the course of an interaction rather than presented all at once, and it is rare for a shared display to differ in perceptual accessibility \cite{HannaTanenhausTrueswell03_CommonGroundPerspective}. Third, none of the experiments reported by Keysar and colleagues included a key comparison condition where the critical item (e.g. the largest measuring cup) was \emph{also} in common ground \cite{BrownSchmidtHanna11_IncrementalPerspectiveTaking}---perhaps people would refer to the critical item less in the privileged condition than in the full common ground condition, which would put a more positive spin on the results. %Note that none of these criticisms invalidate the paradoxical result that listeners move an occluded object, they just suggest an alternative explanation. Reference disambiguation is probabilistic: common ground and goodness of fit are weighed against one another.

In this paper, we offer an additional factor that helps account for Keysar's results. 
Theory of mind as applied within language understanding depends on an accurate model of what a speaker would say in different situations; given an utterance, a listener can then reason backward to the most plausible situation.
This suggests that we consider whether the utterances used by the confederate in Keysar's critical conditions were actually what a speaker would be expected to say. 
If not, then perhaps participants are making choices that are in fact consistent with correct pragmatic interpretation on the confederate's (uncooperative) utterance.
More precisely, when both players know that objects are occluded in the display, the speaker may tend to add additional precision to references in order to avoid confusion. If the listener \emph{expects} the speaker to do this, they will pragmatically pick the \emph{a priori} more likely referent of the referring expression, which in critical trials will be the occluded object. In other words, it is precisely \emph{because} the listener takes the speaker's mental state into consideration that they are tricked by an uncooperative confederate into choosing the wrong item. 
%=======
%Some objects were occluded such that only the matcher could see them, creating an asymmetry in the players' knowledge. To perform accurately, the matcher would need to apply theory of mind and reason about the director's beliefs. For example, if a much larger measuring cup were placed in an occluded slot and agents failed to account for the director's beliefs, they might interpret `the large cup' to mean this occluded cup even though the director couldn't possibly know about it. Indeed, \citeA{KeysarLinBarr03_LimitsOnTheoryOfMindUse} found that participants attempted to move the hidden item in 30\% of cases: 71\% of participants attempted to move this hidden item at least once (out of four critical cases) in the experimental condition, compared to 0\% in a control condition where there was no ambiguity over the referent. 46\% of participants reached for the hidden object at least twice. Additionally, eye-tracking data showed that participants considered the hidden item more often and for longer in the experimental condition than the control condition. 
%>>>>>>> 179ee1de2933b596af109cc10f157c31fdcea761

We began by replicating \citeA{KeysarLinBarr03_LimitsOnTheoryOfMindUse} in a multi-player web experiment. We recruited participants to be both director and matcher (instead of using a confederate), but instructions for critical items, as well as a random subset of filler items, remained scripted as in the original study. We replicated the original finding, but noted a tendency of directors to be ove-rinformative in unscripted trials. We then ran the same experiment without scripting the directions for critical trials, observing unconstrained director behavior. We observe much greater precision in unconstrained director utterances and better performance of the matchers. This minimal pair of experiments demonstrates that listener mistakes are at least partially due to the pragmatics of the task, ironically showing that apparent failures of theory of mind are in fact attributable to sophisticated expectatuions about speaker behavior---that is, to theory of mind.

\ndg{intro can probably be compressed some if needed.}

\begin{table*}
\begin{center}
\begin{tabular}{ p{2cm} | r | r |  r || r | r | r || r | r | r}
& \multicolumn{3}{c||}{\% attempted at least once} & \multicolumn{3}{c||}{\% attempted at least twice} & \multicolumn{3}{c}{\% of total cases}\\
\hline
& Orig. & Exp. 1 & Exp. 2 & Orig. & Exp. 1 & Exp. 2 & Orig. & Exp. 1 & Exp. 2  \\
Experimental & 71 & 93 & 61 & 46 & 57 & 32 & 30 & 43 & 24\\
Baseline        & 0   & 7   & 0   & 0   & 0   & 0   & 0   & 2   & 0\\
\end{tabular}
\caption{Side-by-side comparison of error rates in \citeA{KeysarLinBarr03_LimitsOnTheoryOfMindUse} and our two replications. The first section shows the percentage of \emph{participants} attempting to move the occluded distractor at least one of the four possible cases. The second section shows the percentage attempting to move it at least twice. The third section shows the percentage of \emph{all experimental trials} that the participant actually tried to move the occluded object. \todo[inline]{Big caption maybe unnecessary. Anything else to include in this table? Maybe Keysar et al, 2000, which reported \% of total cases, or our Exp. 2 results if you remove the two most difficult items? There, errors drop quite a lot.}}
\label{table:mainResults}
\end{center}
\end{table*}

@@

\end{tabular}
\caption{Item-wise error rates}
\label{table:ItemWise}
\end{center}
\end{table*}

@@

Removing constraints on the speaker produced a dramatic reduction in overall matcher errors and systematic variation in error rates across items. We found that unscripted speakers systematically tended to produce more precise, costly referring expressions than their scripted counterparts -- for example, ``the clear audiocassette'' when there was only one ``tape'' in the speaker's display. These precise labels were less likely to fit the hidden distractor, and therefore less likely to lead to a matcher error.

We noticed several differences between our study and \citeA{KeysarLinBarr03_LimitsOnTheoryOfMindUse} that prevented a direct replication. The primary difference between our study and the original is the fact that it will be run online with participants connected via a virtual environment, instead of face-to-face in a room. A related difference is our decision not to use a confederate. In general, studies using confederates are difficult to replicate exactly and are impractical to run online. We are aware that this change (while keeping scripted instructions, which were the primary justification for the confederate) may lead to additional variation across groups. It is also known that textual communication (via our instant messaging interface) can differ from verbal communication (as in the original study). Finally, the way people attempt to move things in our virtual environment may be different from how people attempt to move things in the lab, and we cannot eye-tracking data to back up our analysis, as the original paper did.

% Second, it's worth thinking about some ways our online version may differ from the original. The fact that we didn't use a confederate didn't seem to be a problem, given our use of randomly scripted instructions. Similarly, because players could type messages directly to one another, and since the director could watch the matcher moving objects in real-time, it's also fair to say that participants truly believed they were playing with another human being.

\ndg{mention overinformativity in refexp somewhere in here...}

One major concern is our graphical representation of occluded cells. We told players that items behind black cells were hidden to the director, showed them side-by-side displays of the director and matcher views, and quizzed them on this detail in the short attention check that players had to pass before playing. But they may have forgotten part-way through the experiment, or failed to fully internalize it. In the in-lab version, there was an actual divider between players, so it would have been more salient. We suspect that some subset of errors in our version game aren't due to a failure of perspective-taking, but instead because of a misunderstanding of what the other person's perspective is to begin with. For example, one matcher wrote, ``This seems to work wether [sic] the space is black or not.'' This suggests that they thought that the black squares were obstacles or invalid destinations. However, this source of errors is held constant across our two experiments, and it would only bias our estimates of errors toward supporting Keysar's results, making our analysis conservative.

At a methodological level, the pair of experiments we report is a reminder that manipulations administered by a trained confederate can interact in unexpected ways with a participant's social and communicative expectations. Regardless of the experimental context, it's illuminating to see how real participants naturally interact, and

\todo[inline]{rdh: methodological note that running multi-player control w/o confederate is important to know whether confederate behavior is valid and cooperative}

>>>>

\begin{figure*}[t!]
\begin{center}
\includegraphics[scale = .5]{images/overall_view.png}
\end{center}
\vspace{-.25cm}
\caption{Interface used in the reported experiments. Objects behind the black squares were hidden from the director.}
\label{fig:interface}
\end{figure*}

@@

In this paper we argue that the apparent failures used to support this view are in fact successes for sophisticated social reasoning. In particular, we argue that critical utterances used by Keysar and colleagues are \emph{uncooperative}: they are less informative than what a speaker would actually produce in that situation; listeners who are sensitive to the pragmatics of the situation expect these more informative utterances and produce ``errors'' when their expectations are flouted.

The argument of Keysar and colleagues is based on an elegant experimental paradigm, where participants played a simple communication game with a confederate. The two players were placed on opposite sides of a $4 \times 4$ grid containing a set of everyday objects (see Fig. \ref{fig:interface}). The confederate played the role of `director,' giving instructions about how to move objects around a grid, and the participant played the role of `matcher,' attempting to follow these instructions. For example, the objects in one trial included a cassette tape. The director gave an instruction like `move the tape up one square,' referring to the cassette.
Critically, some objects were occluded such that only the matcher could see them, creating an asymmetry in the players' knowledge. To perform accurately on critical trials, the matcher would need to apply theory of mind to reason about the director's beliefs. For example, imagine a much larger measuring cup were placed in an occluded slot: if a participant failed to account for the director's (partial) knowledge, she might interpret `the large cup' to mean this occluded cup (which the director couldn't possibly know about). 
Indeed, \citeA{KeysarLinBarr03_LimitsOnTheoryOfMindUse} found that participants attempted to move the hidden item in 30\% of cases: 71\% of participants attempted to move this hidden item at least once (out of four critical cases) in the experimental condition, compared to 0\% in a control condition where there was no ambiguity over the referent. 
Additionally, eye-tracking data showed that participants considered the hidden item more often and for longer in the experimental condition than the control condition.

While these results are compelling, the paradigm has been criticized from several different angles. First, \citeA{HellerGrodnerTanenhaus08_Perspective} has pointed out that in many cases, the hidden object was a better fit for the referring expression than the one in common ground (e.g. the hidden roll of tape vs. the cassette tape for ``the tape''), making the hidden object \emph{a priori} more likely to be the referent. Although relative fit has not been empirically evaluated, it would be fairer to compare two objects that fit the referring expression equally well. Second, the viewpoint asymmetry paradigm is somewhat unnatural: common ground is typically built incrementally over the course of an interaction rather than presented all at once, and it is rare for a shared display to differ in perceptual accessibility \cite{HannaTanenhausTrueswell03_CommonGroundPerspective}. Third, none of the experiments reported by Keysar and colleagues included a key comparison condition where the critical item (e.g. the largest measuring cup) was \emph{also} in common ground \cite{BrownSchmidtHanna11_IncrementalPerspectiveTaking}---perhaps people would refer to the critical item less in the privileged condition than in the full common ground condition, which would put a more positive spin on the results. %Note that none of these criticisms invalidate the paradoxical result that listeners move an occluded object, they just suggest an alternative explanation. Reference disambiguation is probabilistic: common ground and goodness of fit are weighed against one another.

In this paper, we offer an additional factor that helps account for Keysar's results. 
Theory of mind as applied within language understanding depends on an accurate model of what a speaker would say in different situations; given an utterance, a listener can then reason backward to the most plausible situation \cite{FrankGoodman12_PragmaticReasoningLanguageGames, GoodmanStuhlmuller13_KnowledgeImplicature}.
This suggests that we consider whether the utterances produced by the confederate in Keysar's critical conditions were actually what a speaker in that context would be expected to say. 
If not, then perhaps listeners are making choices that are in fact consistent with a correct pragmatic interpretation of the confederate's (uncooperative) utterance.
More precisely, when both players know that objects are occluded in the display, the speaker may tend to add additional precision to references in order to avoid confusion. If the listener \emph{expects} the speaker to do this, they will pragmatically pick the \emph{a priori} more likely referent of the referring expression, which in critical trials will be the occluded object. In other words, it is precisely \emph{because} the listener takes the speaker's mental state into consideration that they are tricked by an uncooperative confederate into choosing the wrong item.

We began by replicating \citeA{KeysarLinBarr03_LimitsOnTheoryOfMindUse} in a multi-player web experiment. We recruited participants to be both director and matcher (instead of using a confederate), but instructions for critical items, as well as a random subset of filler items, remained scripted as in the original study. We replicated the original finding, but noted a tendency of directors to be overinformative in unscripted trials. We then ran the same experiment without scripting the directions for critical trials, observing unconstrained director behavior. We observe much greater precision in unconstrained director utterances, which match targets much better than distractors, and better performance of the matchers. This minimal pair of experiments demonstrates that listener mistakes are at least partially due to the pragmatics of the task, ironically showing that apparent failures of theory of mind are in fact attributable to sophisticated expectations about speaker behavior---that is, to theory of mind.

\begin{table*}
\begin{center}
\begin{tabular}{ p{2cm} | r | r |  r || r | r | r || r | r | r}
& \multicolumn{3}{c||}{\% attempted at least once} & \multicolumn{3}{c||}{\% attempted at least twice} & \multicolumn{3}{c}{\% of total cases}\\
\hline
& Orig. & Exp. 1 & Exp. 2 & Orig. & Exp. 1 & Exp. 2 & Orig. & Exp. 1 & Exp. 2  \\
Experimental & 71 & 93 & 61 & 46 & 57 & 32 & 30 & 43 & 24\\
Baseline        & 0   & 7   & 0   & 0   & 0   & 0   & 0   & 2   & 0\\
\end{tabular}
\caption{Side-by-side comparison of error rates in \citeA{KeysarLinBarr03_LimitsOnTheoryOfMindUse} and our two replications. The first section shows the percentage of \emph{participants} attempting to move the occluded distractor at least one of the four possible cases. The second section shows the percentage attempting to move it at least twice. The third section shows the percentage of \emph{all experimental trials} that the participant actually tried to move the occluded object. }
\label{table:mainResults}
\end{center}
\end{table*}

@@

\end{tabular}
\caption{Item-wise error rates for critical trials}
\label{table:ItemWise}
\end{center}
\end{table*}

@@

Pragmatic language understanding requires sophisticated social reasoning. To interpret an utterance, a listener must consider the speaker's communicative goals and context. In both experiments, we found evidence of such considerations: when the speaker's utterance fit the hidden distractor better than the target, matchers were more likely to make errors. This suggests that reference disambiguation was driven more by \emph{a priori} label fitness than consideration of the speaker's epistemic state, as Keysar and colleagues claim.

However, this does not necessarily imply limitations on theory of mind. If listeners arrive at this interpretation strategy through pragmatic reasoning, then neglecting the speaker's epistemic state would be rational and also the result of theory of mind use. In Exp.~2, we found that such pragmatic interpretations were in fact justified: speakers naturally produce more precise, informative utterances than required and these unconstrained utterances fit the target significantly better than they fit the hidden distractor. Thus, the patterns of errors observed in Exp.~1 can be explained as an uncooperative confederate setting up certain pragmatic expectations through the task context, then deliberately flouting them in critical trials. % Removing constraints on the speaker produced a dramatic reduction in overall matcher errors and systematic variation in error rates across items. We found that unscripted speakers systematically tended to produce more precise, costly referring expressions than their scripted counterparts -- for example, ``the clear audiocassette'' when there was only one ``tape'' in the speaker's display. These precise labels were less likely to fit the hidden distractor, and therefore less likely to lead to a matcher error.

It is worth noting several significant differences between our study and \citeA{KeysarLinBarr03_LimitsOnTheoryOfMindUse} that prevented a direct replication. The primary difference between our study and the original, of course, is that it was run on the web with participants connected through a virtual environment, instead of face-to-face in a room. We believe we addressed the major concern about exploring theory of mind in web experiments -- that participants do not truly believe they are interacting with another human -- by allowing instantaneous, responsive, real-time interaction.  On the other hand, it is known that textual communication, as in our chat box, can differ from face-to-face verbal communication, and there was no way of addressing this. % Directors watched objects move through the actual moment-by-moment mouse trajectories that matchers used to drag them, and matchers saw the actual timing of the messages sent by the director.

A related difference is our decision not to use a confederate. While a confederate is useful for reducing variation across instances of the experiment, we see several downsides. Beyond the difficulties of conducting large-scale experiments with confederates on the web, it is difficult to exactly replicate all the subtleties of the confederate's behavior that might influence the .  At a methodological level, the pair of experiments we report is a reminder that manipulations administered by a trained confederate can interact in unexpected ways with a participant's social and communicative expectations. Regardless of the experimental context, it's illuminating to see how real participants naturally interact, and

A final concern is our graphical representation of occluded cells. We told players that items behind black cells were hidden to the director, showed them side-by-side displays of the director and matcher views, and quizzed them on this detail in the short attention check that players had to pass before playing. But they may have forgotten part-way through the experiment, or failed to fully internalize it. In the in-lab version, there was an actual divider between players, so it would have been more salient. We suspect that some subset of errors in our version game aren't due to a failure of perspective-taking, but instead because of a misunderstanding of what the other person's perspective is to begin with. However, this source of errors is held constant across our two experiments, and it would only bias our estimates of errors upward.

The Gricean maxim of quality dictates that cooperative speakers should make their utterance as informative as is required for current purposes, but no more so. If the referring expressions used in Exp.~1 uniquely pick out the target from the speaker's perspective, then why would unscripted speakers in Exp.~2 strive to be more informative than strictly necessary and why would listeners come to expect such ``overinformativity'' through social reasoning? Studies of overinformativity in referring expressions have uncovered a number of factors, such as the tendency to mention perceptually salient features to speed up the identification process  \cite{KoolenGattGoudbeekKrahmer11_Overspecification}. In Keysar's task, it seems particularly relevant that both participants knew about the presence of hidden objects, increasing uncertainty about possible distractors and placing additional pressure on the speaker to overspecify.

It is interesting that the item-wise variability we observed in both experiments is at least partially due to the availability and effectiveness of more specific expressions. For example, no speaker in Exp.~2 produced ``the bottom block,'' which was used as the critical instruction in Exp.~1. Instead, they said ``the bluish block with a B'' or ``block with the blue writing,'' which relied on less confusable perceptual features and thereby decreased error rates. The items on which errors were most likely in Exp.~2 were difficult precisely because the overinformative utterance was \emph{still} ambiguous: even if the speaker uses ``hairbrush'' instead of the basic-level ``brush,'' for example, the distractor remains a more typical exemplar of the subcategory.
